Emulating Comparative Oncology Trials with Real-world Evidence Studies (ENCORE): Process Development and Methodological Considerations for Oncology Real-World Data

### 2025-07-06 manuscript/manuscript.tex
Manuscript updated

\KOMAoption{captions}{tableheading,figureheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={A Process for Emulating Comparative Oncology Trials with Real-world Evidence Studies (ENCORE): Process Development and Methodological Considerations for Oncology Real-World Data},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{A Process for Emulating Comparative Oncology Trials with
Real-world Evidence Studies (ENCORE): Process Development and
Methodological Considerations for Oncology Real-World Data}
\author{}
\date{}
\begin{document}
\maketitle

\textbf{Authors}: Janick Weberpals\textsuperscript{1}, Kenneth L.
Kehl\textsuperscript{2}, Donna R. Rivera\textsuperscript{3}, Pallavi
Mishra-Kalyani\textsuperscript{3}, Catherine C.
Lerro\textsuperscript{3}, Erin Larkins\textsuperscript{4}, Sonia
Singh\textsuperscript{4}, Preeti Narayan\textsuperscript{4}, Richard
Curley\textsuperscript{4}, Paul Kluetz\textsuperscript{4}, Georg
Hahn\textsuperscript{1}, Priyanka Anand\textsuperscript{1}, Yanina
Natanzon\textsuperscript{5}, Andrew J. Belli\textsuperscript{6},
Ching-Kun Wang\textsuperscript{6}, Jenna
Collins\textsuperscript{7}\textbf{,} Jonathan
Kish\textsuperscript{7}\textbf{,} Janet Espirito\textsuperscript{8},
Nicholas J Robert\textsuperscript{8}, Robert J.
Glynn\textsuperscript{1}, Sebastian Schneeweiss\textsuperscript{1},
Shirley V. Wang\textsuperscript{1}

@@

@@

\emph{Manuscript last updated: 2025-05-28 17:11:23.084986}

@@

Randomized controlled trials (RCTs) are the go-to methodology for
establishing the efficacy and safety of medical products. Directed by
the 21\textsuperscript{st} Century Cures Act,\textsuperscript{1} the
Food and Drug Administration (FDA) established a framework to factor
consideration of real-world evidence (RWE) generated from routine-care
health data such as electronic health records (EHR) to evaluate and
contextualize the safety and effectiveness of novel cancer
therapies.\textsuperscript{2} Accounting for 30\% of new drug approvals,
oncology was the disease area with the most FDA drug approvals in
2024\textsuperscript{3} and has many areas of high unmet medical need to
treat serious conditions. Therefore, RWE has particularly important
potential to complement evidence from RCTs in the field of oncology.
Potential use cases include the assessment of effectiveness in specific
patient populations that are underrepresented in RCTs, the construction
of external control arms in single-arm trials that have feasibility or
equipoise challenges, or the precision oncology-focused discovery of
biomarkers among pan-tumor populations that harbor specific genomic and
immuno-pathological signatures.

However, the validity and transportability of results derived between
RWE studies and RCTs depends on multiple factors. Frequently referenced
limitations include lack of baseline randomization, missing data,
unmeasured confounding, small study sizes, data
discontinuity,\textsuperscript{4, 5} adoption of changes in guidelines
in real-world care and the inability to measure and emulate common
eligibility criteria, including prognostic factors, and standardized
response assessments in real-world data (RWD).\textsuperscript{6} While
examples of oncology trial emulations have been
published,\textsuperscript{6--8} a systematic and scaled approach to
emulate a diverse set of oncology trials with multiple heterogeneous
databases is necessary to gain confidence in the validity of RWE
studies, evaluate regulatory considerations, and to provide context as
to which questions can be validly answered.

The RCT DUPLICATE initiative\textsuperscript{9} increased our
understanding of when RWE studies can come to causal conclusions on
treatment effects by benchmarking results against RCTs under the
assumption that each RCT finding reflects a causal treatment effect. In
settings where the RCT designs could be emulated well, RWE studies were
able to reach similar conclusions.\textsuperscript{10} However, this
prior work from RCT-DUPLICATE focused primarily on emulating trials in
non-oncology settings using claims databases.

The \emph{Emulation of Comparative Oncology Trials with Real-world
Evidence} (ENCORE) project\textsuperscript{11} aims to extend this work
to oncology. Studies in oncology come with their own unique set of
methodological challenges which must be systematically explored and
understood. Building on a process co-developed with the FDA through RCT
DUPLICATE,\textsuperscript{9} this expansion to oncology will emulate 12
randomized clinical trials using multiple specialty oncology EHR data
sources. The process will emphasize transparency and include documented
data fit-for-purpose assessment\textsuperscript{12, 13} for each RWD
source with respect to each trial emulation as well as extensive
sensitivity analyses to assess robustness of findings.

The objectives of this project are to 1) develop state-of-the-art
methodological approaches and 2) to apply them to create insights into
the potential use of RWE to enhance regulatory science in oncology. To
achieve these objectives, this demonstration project will systematically
emulate 12 oncology trials across four cancer types and assess the
agreement of treatment effect estimates between RCTs and their
respective emulations.

Here, we describe the design and process for the selection of the 12
oncology RCTs, the assessment of the database quality and selection,
protocol development and pre-registration, study design and statistical
analysis, and pre-specified agreement metrics to evaluate the
concordance between RCTs and emulations.

@@

The focus of ENCORE is to efficiently explore in what contexts RWE
studies can or cannot yield similar results compared to RCTs. Therefore,
the project emphasizes trials of therapies for common cancers for which
there has been substantial therapeutic development in recent years.
After review and collaboration with clinical and regulatory experts,
four cancer indications were identified: non-small cell lung cancer,
breast cancer, colorectal cancer and multiple myeloma. For each cancer
type, we aim to conduct three trial emulations using multiple accessible
databases (i.e., the total equaling 12 trial emulations x \emph{n}
databases which are found fit-for-purpose for each trial).

We used a systematic process for trial selection where the eligibility
criteria are documented in CONSORT diagrams showing reasons for
excluding RCTs for each cancer type. The search was conducted using the
\emph{Aggregated Analysis of ClinicalTrials.gov} (AACT) database which
is a publicly available relational database developed and maintained by
the Clinical Trials Transformation Initiative (CTTI) that contains
information (protocol and result data elements) about studies registered
on ClinicalTrials.gov.\textsuperscript{14} To identify eligible trials,
we used a combined search query strategy of the National Library of
Medicine (NLM)-controlled \emph{Medical Subject Headings
\textbf{(}}MeSH) term and a free keyword search for the respective
cancer indication in the \emph{conditions}, \emph{studies} and
\emph{detailed\_descriptions} fields of each trial entry on
ClinicalTrials.gov.

@@

\begin{itemize}
\item
  Interventional
\item
  Randomized
\item
  Intervention model: parallel assignment
\item
  Industry-sponsored
\item
  Trial start in 2011 or later
\item
  Primary purpose was to study treatment efficacy
\item
  Overall survival must be one of the endpoints reported (either as
  hazard ratio or median overall survival time)
\item
  Recruitment status: `Completed' or `Active, not recruiting'
\item
  Feasibility and clinical relevance (latter was defined as treatment or
  paradigm-changing trials or such that challenged existing treatment
  policies)
\end{itemize}

The rationale and operationalization for each criterion is listed in
detail in Table~\ref{tbl-criteria}. We will mainly consider pivotal
large late-phase interventional RCTs after 2011 because treatment
guidelines among included cancer indications have undergone significant
changes in recent years. Due to the rapid adoption of breakthrough
therapies in routine care, few patients currently initiate outdated
treatment regimens in current clinical practice. Conversely, we excluded
trials with results published too recently to allow for enough data and
follow-up time accrual in the real-world databases used for this
project. We did not define a global cut-off as the requirements for
follow-up time are different for each cancer type and population (e.g.,
advanced NSCLC versus early-stage breast cancer) and the decisions were
made on a case-by-case basis.

Although there have been substantial methodological advancements to
increase our understanding on the emulation and comparison of real-world
progression-free survival (PFS) and objective response rates (ORR) to a
RECISTv1.1\textsuperscript{15}-based PFS and ORR assessment in
RCTs\textsuperscript{16, 17}, imaging-based evaluations still hold a
level of granularity which is not well reflected in chart-abstracted
assessments of a patient's progression in routine
care.\textsuperscript{18, 19} The timing and cadence of intervals
between progression assessments can differ between RCTs and routine care
which may result in measurement error and bias.\textsuperscript{20}
Given this, and the large number of other methodological challenges
including missing data, small sample sizes, data discontinuity and
rapidly changing guideline treatments, we focus on the emulation of
overall survival (OS) as the endpoint of interest. Therefore, we include
only trials that have reported overall survival (OS) as one of the
pre-specified endpoints in the protocol.

While most trial eligibility criteria could be operationalized in an
automated fashion, our final criterion was an assessment of emulation
feasibility and clinical relevance. This criterion involved extensive
human review. The fundamental points considered in this step include an
initial feasibility assessment of the data fitness,\textsuperscript{19}
involving an assessment of whether critical eligibility criteria (e.g.,
biomarker status) including prognostic factors (e.g., ECOG performance
score) are measurable, and whether preliminary study size based on a
rough estimation of the number of patients observed in the data with the
combination of treatment regimen and line of treatment is reasonable for
shortlisting the trial. Lastly, trial candidates were ranked and
shortlisted into primary and runner-up candidates based on their
clinical and regulatory relevance.

A list of tentative, shortlisted primary candidates is presented in
Table~\ref{tbl-rcts}. The corresponding selection process is illustrated
in the CONSORT diagrams (\textbf{Supplementary Figures 1-4}) and runner
up candidate trials are listed in \textbf{Supplementary Table 1}.
Naturally, the majority of shortlisted trials cover advanced (locally
advanced, inoperable, recurrent/progressive disease) or metastatic
cancer populations because a large proportion of drug development
efforts have recently focused on these settings. A key objective that we
aim to explore with the selected trials is to achieve a better
understanding how different disease settings (early, late), line
settings ({[}neo{]}adjuvant, first line, advanced lines of therapy),
therapy protocols (monotherapy, combination therapy) and clinical
characteristics (all-comer versus complex genetic or immunological
signatures) can be emulated using RWD. If more thorough feasibility
assessments after applying all relevant eligibility criteria suggest
that the threat of bias from mis-measurement of key study parameters or
residual confounding remains high, or that the study size is not
sufficient, then runner-up candidate trials will be considered instead.

@@

The ENCORE project will use data from four oncology-specific electronic
health records (EHR)-derived RWD data sources (in alphabetical order):
ConcertAI, COTA, Flatiron Health, and Ontada/McKesson. The included
databases comprise the necessary information to study medical product
effectiveness in oncology derived from US-based oncology EHRs. A
detailed description and sampling methodology will be provided with each
trial emulation protocol. For ENCORE, not all databases will be
available for each cancer indication and the names of the databases will
be blinded and referred to as ENCORE DataBase (EDB) 1, 2, 3 and 4 for
the final reporting of results (the numbering does not coincide with the
above order of mention of the databases). If more than one database is
considered fit-for-purpose for a respective trial emulation, the best
possible analytic model will be employed for each database separately
and final treatment effect estimates will be pooled using a
meta-analytic approach using fixed effects (weights reflect the inverse
of the variance of each database study) and random effects (weights
reflect the inverse of the variance of each database study plus an
additional variance term that quantifies the assumed heterogeneity
between databases) models.\textsuperscript{21}

@@

For each selected RCT, a detailed protocol, pre-specifying key elements
of the trial emulation, will be developed using the HARPER protocol
template\textsuperscript{22} recommended for regulatory submissions of
RWE studies\textsuperscript{23} and will be registered on
ClinicalTrials.gov after review by a clinical and FDA regulatory expert
panel. Following the target trial emulation framework, we will provide
an explicit statement and rationale on how each element will be emulated
including database selection, covariate measurement, operationalization
of key eligibility criteria, study design, data analysis and causal
contrasts of interest.\textsuperscript{24, 25} Since it is common that
oncology RCTs update OS estimates periodically based on accrued
follow-up time, the protocol will give a brief summary of each emulated
RCT and specify which target OS estimates will be used for agreement
metric evaluation (see Section~\ref{sec-agreement-metrics}). All
eligibility criteria will be extracted based on publications, publicly
available protocols, and statistical analysis plans of the selected RCT.

@@

One important aspect when emulating oncology trials is the choice and
estimation of the appropriate estimand of interest.\textsuperscript{26}
In prior RCT-DUPLICATE efforts, a ``while on treatment'' strategy was
chosen for database studies that were designed to emulate trials where
the primary trial analysis was intention-to-treat (e.g., a ``treatment
policy'' estimand). This estimand was chosen to mimic the high adherence
typically observed in large cardiovascular clinical trials in the
context of low adherence in clinical practice. However, a ``while on
treatment'' estimand would be inappropriate when emulating oncology
trials, not only because of highly informative censoring, but also
because OS outcome analyses in oncology are typically ``treatment
policy'' estimands that allow crossover.\textsuperscript{27} We plan to
focus on ``treatment policy'' estimands for the primary analyses in our
trial emulations. However, allowing crossover can dilute treatment
effects, moving point estimates toward the null. Therefore, differences
in cross-over and discontinuation rates between the trial and emulation
may still contribute to estimand differences. Recognizing these
practical challenges, we will carefully characterize the patterns of
discontinuation or cross-over in both the trials and the database
studies designed to emulate them.

@@

Real-world data fitness and emulation feasibility for each shortlisted
candidate trial will be assessed in multiple steps based on direction
from of the oncology quality, characterization, and assessment of
real-world data (Oncology QCARD) Initiative.\textsuperscript{19} The
first step assesses if relevant variables like exposure, line of
therapy, outcomes, and covariates are generally available, measured and
operationalizable in routine-care. Since many oncological RCTs in recent
years have focused on selected, biomarker-defined populations, nuances
in measurement and operationalizability of specific biomarkers must be
reflected to ensure a representative and large enough study population.
For example, immune checkpoint inhibitors have significantly changed the
cancer treatment landscape since the first approvals of the CTLA-4
inhibitor ipilimumab in 2011 and the PD-1 inhibitors pembrolizumab and
nivolumab in 2014 in the United States. However, the operationalization
of the expression of the PD-L1 biomarker in RCTs (e.g., as a percent
staining, tumor proportion score or combined positive score) has
evolved, and as such PD-L1 '\emph{positivity'} may have different
definitions across calendar years based on different cut-off values.

According to the \emph{Structured Process to Identify Fit-For-Purpose
Data} (SPIFD) framework\textsuperscript{13}, the next step outlines how
eligibility criteria will be ascertained using a color-coded heatmap
that will indicate the level of confidence on how well each criterion
can be emulated in each selected database. As there are general
eligibility criteria in oncology clinical trials which either are
infeasible to emulate (e.g., physician-assessed survival prognosis of x
months) or that are clinically not relevant for the emulation of the
trial (e.g., male patients should be willing to use barrier
contraception), the study team will decide on key eligibility criteria
for the emulation of the trial.

Additionally, we will provide conceptual and operationalized definitions
on how exposures, outcomes and covariates are defined in each respective
database. There will be special emphasis on how exposure, in context of
disease and line of therapy settings, and the OS outcome are emulated.
For all considered databases, OS is typically a composite measure that,
depending on the underlying database, can be derived from different
mortality sources (i.e., EHR documentation, social security death index,
obituary and other linkages). Given that not all relevant sources that
provide mortality data are synchronized and updated uniformly,
sensitivity analyses with more conservative (i.e., earlier) censoring
dates will be considered for each trial emulation to mitigate the
potential impact of ghost-time bias.\textsuperscript{28}

@@

Particularly when emulating pivotal trials of paradigm-changing
treatments, multiple aspects need to be considered, such as the
contemporaneity of the control cohort, the adoption rate of the novel
medical product in routine care, the magnitude of the clinical treatment
benefit and the rate in which patients discontinue or cross-over, which
could dilute effect estimates.

To that end, comprehensive data explorations will be performed as part
of the protocol development to contextualize these parameters and (if
reported) draw comparisons to the emulated trial. Examples for such
standard diagnostics are visualized in Figure~\ref{fig-initiators}. All
exploratory analyses will be conducted blinded towards the outcome to
avoid influencing study design and analytic choices.

The distribution of patient characteristics, by exposure status, will be
examined using Table 1's before and after applying eligibility criteria
and contrasted with the distributions of patient characteristics of the
original RCT. Initial propensity score matching or weighting methods
will be applied to ensure that measured pre-exposure covariates can be
balanced, exposure cohorts are conditionally exchangeable at baseline
and resulting sample sizes are still sufficient after matching or
weighting. At this stage, all exploratory analyses will be conducted
blinded towards the outcome to not bias any study design or analytic
choices based on known outcome information.

>>>>

\KOMAoption{captions}{tableheading,figureheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Emulating Comparative Oncology Trials with Real-world Evidence Studies (ENCORE): Process Development and Methodological Considerations for Oncology Real-World Data},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Emulating Comparative Oncology Trials with Real-world Evidence
Studies (ENCORE): Process Development and Methodological Considerations
for Oncology Real-World Data}
\author{}
\date{}
\begin{document}
\maketitle

\textbf{Authors}: Janick Weberpals\textsuperscript{1}, Sebastian
Schneeweiss\textsuperscript{1}, Kenneth L. Kehl\textsuperscript{2},
Donna R. Rivera\textsuperscript{3}, Pallavi
Mishra-Kalyani\textsuperscript{3}, Catherine C.
Lerro\textsuperscript{3}, Erin Larkins\textsuperscript{4}, Sonia
Singh\textsuperscript{4*}, Preeti Narayan\textsuperscript{4}, Richard
Curley\textsuperscript{4}, Georg Hahn\textsuperscript{1}, Priyanka
Anand\textsuperscript{1}, Yanina Natanzon\textsuperscript{5}, Andrew J.
Belli\textsuperscript{6}, Ching-Kun Wang\textsuperscript{6}, Jenna
Collins\textsuperscript{7}\textbf{,} Jonathan
Kish\textsuperscript{7}\textbf{,} Janet Espirito\textsuperscript{8},
Nicholas J Robert\textsuperscript{8}, Robert J.
Glynn\textsuperscript{1}, Shirley V. Wang\textsuperscript{1}

@@

*At the time this research was conducted

@@

\emph{Manuscript last updated: 2025-07-06 14:54:57.141696}

@@

Randomized controlled trials (RCTs) are the evidentiary gold standard
methodology for establishing the efficacy and safety of medical
products. Guided by the 21\textsuperscript{st} Century Cures
Act,\textsuperscript{1} the Food and Drug Administration (FDA)
established a framework to factor considerations for use of real-world
evidence (RWE) generated from real-world data (RWD) such as electronic
health records (EHR) to support approval of new indications or satisfy
postmarketing requirements.\textsuperscript{1} Accounting for
approximately 30\% of new FDA drug approvals, oncology was the disease
area with the most new approvals in 2024\textsuperscript{2} as well as
several indication expansions, and has many areas of high unmet medical
need to treat serious conditions. Therefore, RWE can have a particularly
important potential to complement evidence from RCTs in the field of
oncology. Potential uses include the assessment of effectiveness in
specific patient populations that are not adequately represented in
RCTs, or the precision oncology-focused discovery of biomarkers among
pan-tumor populations that harbor specific genomic and
immuno-pathological signatures.

However, the comparability and transportability of results derived
between observational studies and RCTs depends on multiple factors.
Frequently referenced limitations include lack of baseline randomization
and the imbalance in prognostic factors resulting thereof, missing data,
unmeasured confounding, small study sizes, data
discontinuity,\textsuperscript{3, 4} adoption of changes in treatment
guidelines over time in routine clinical care and the inability to
measure and emulate common eligibility criteria, including prognostic
factors, and standardized response assessments in real-world data
(RWD).\textsuperscript{5} While examples of oncology trial emulations
have been published,\textsuperscript{5--7} a systematic and scaled
approach to emulate a diverse set of oncology trials with multiple
heterogeneous databases is necessary to increase confidence in the
interpretability of observational studies, evaluate regulatory
considerations, and to provide context as to which questions can be
validly answered.

The RCT DUPLICATE initiative\textsuperscript{8} increased our
understanding of when observational studies can come to causal
conclusions on treatment effects by benchmarking results against RCTs
under the assumption that each RCT finding reflects a causal treatment
effect. In settings where the RCT designs could be emulated well,
observational studies were able to reach similar
conclusions.\textsuperscript{9} However, this prior work from
RCT-DUPLICATE focused primarily on emulating trials in non-oncology
settings using claims databases.

The \emph{Emulation of Comparative Oncology Trials with Real-world
Evidence} (ENCORE) project\textsuperscript{10} aims to extend this work
to oncology. Clinical studies in oncology come with unique
methodological challenges which must be systematically explored and
understood. Building on a process co-developed with the FDA through RCT
DUPLICATE,\textsuperscript{8} this expansion to oncology will emulate 12
randomized clinical trials using multiple specialty oncology EHR data
sources. The process will emphasize transparency and include a data
fitness-for-use assessment\textsuperscript{11, 12} for each RWD source
with respect to each trial emulation as well as extensive sensitivity
analyses to assess robustness of findings.

The objectives of this project are to 1) develop state-of-the-art
methodological approaches and 2) apply these methods to gain insights
into the potential use of RWE to enhance regulatory and clinical
decision-making in oncology. To achieve these objectives, this
demonstration project will systematically emulate 12 oncology trials
across four cancer types and assess the agreement of treatment effect
estimates between RCTs and their respective emulations.

Here, we describe the design and process for the selection of the 12
oncology RCTs, assessment of the database quality and selection,
protocol development and pre-registration, study design and statistical
analysis plans, and pre-specified agreement metrics to evaluate the
concordance between RCTs and emulations.

@@

The focus of ENCORE is to efficiently explore in what contexts
observational studies can or cannot yield similar results compared to
oncology RCTs. Therefore, the project emphasizes trials of therapies for
common cancers for which there has been substantial therapeutic
development in recent years. After review and collaboration with
clinical and regulatory experts, four cancer indications were
identified: non-small cell lung cancer, breast cancer, colorectal cancer
and multiple myeloma. For each cancer type, we aim to conduct three
trial emulations using multiple accessible databases (i.e., the total
equaling 12 trial emulations x \emph{n} databases which are found
fit-for-purpose for each trial).

We used a systematic process for trial selection where the eligibility
criteria are documented in CONSORT diagrams showing reasons for
excluding RCTs by cancer type. The search was conducted using the
\emph{Aggregated Analysis of ClinicalTrials.gov} (AACT) database which
is a publicly available relational database developed and maintained by
the Clinical Trials Transformation Initiative (CTTI) that contains
information (protocol and result data elements) about studies registered
on ClinicalTrials.gov.\textsuperscript{13} To identify eligible trials,
we used a combined search query strategy of the National Library of
Medicine (NLM)-controlled \emph{Medical Subject Headings
\textbf{(}}MeSH) term and a free keyword search for the respective
cancer indication in the \emph{conditions}, \emph{studies} and
\emph{detailed\_descriptions} fields of each trial entry on
ClinicalTrials.gov.

@@

\begin{itemize}
\item
  Interventional
\item
  Randomized
\item
  Intervention model: Parallel assignment
\item
  Industry-sponsored
\item
  Trial start: 2011 or later
\item
  Primary purpose: Study treatment efficacy
\item
  Endpoint(s): Overall survival must be one of the endpoints reported
  (either as hazard ratio or median overall survival time)
\item
  Recruitment status: `Completed' or `Active, not recruiting'
\item
  Feasibility and clinical relevance (latter was defined as treatment or
  paradigm-changing trials or such that challenged existing treatment
  policies)
\end{itemize}

The rationale and operationalization for each criterion is listed in
detail in Table~\ref{tbl-criteria}. We mainly considered pivotal large
late-phase RCTs that were initiated after 2011 because treatment
guidelines for included cancer indications have undergone significant
changes in recent years. Due to the rapid adoption of newly-approved
therapies in routine care, patients are less likely to receive outdated
treatment regimens in clinical practice. Conversely, we excluded trials
with results published too recently to allow for substantive medical
product uptake and follow-up time accrual in real-world databases used
for this project. We did not define a global cut-off as the requirements
for follow-up time are different for each cancer type and population
(e.g., advanced NSCLC versus early-stage breast cancer) and the
decisions were made on a clinically relevant basis.

Although there have been substantial methodological advancements to
increase our understanding on the emulation and comparison of real-world
progression-free survival (PFS) and objective response rates (ORR) to a
RECISTv1.1\textsuperscript{14}-based PFS and ORR assessment in
RCTs\textsuperscript{15, 16}, imaging-based evaluations still require a
level of granularity which is not well reflected in chart-abstracted
assessments of a patient's progression in routine
care.\textsuperscript{17, 18} The timing and cadence of intervals
between progression assessments can differ between RCTs and routine care
which may result in measurement error and bias.\textsuperscript{19}
Given this, and the large number of other methodological challenges
including missing data, small sample sizes, data discontinuity and
rapidly changing guideline treatments, we focus on the emulation of
overall survival (OS) as the endpoint of interest. Therefore, we include
only trials that have reported OS as a pre-specified endpoint in the
protocol.

While most trial eligibility criteria could be operationalized in an
automated fashion, our final criterion was an assessment of emulation
feasibility and clinical relevance. This criterion involved extensive
human review. The fundamental points considered in this step include an
initial feasibility assessment of the data fitness,\textsuperscript{18}
involving an assessment of whether critical eligibility criteria (e.g.,
biomarker status) including prognostic factors (e.g., Eastern
Cooperative Oncology Group {[}ECOG{]} performance score) are measurable,
and whether preliminary study size based on a rough estimation of the
number of patients observed in the data with the combination of
treatment regimen and line of therapy is reasonable for shortlisting the
trial. Lastly, trial candidates were ranked and shortlisted as primary
or runner-up based on their clinical and regulatory relevance.

A list of tentative, shortlisted primary candidates is presented in
Table~\ref{tbl-rcts}. The corresponding selection process is illustrated
in the CONSORT diagrams (\textbf{Supplementary Figures 1-4}) and runner
up candidate trials are listed in \textbf{Supplementary Table 1}. The
majority of shortlisted trials represent indications for advanced
(locally advanced, inoperable, recurrent/progressive disease) or
metastatic cancer patients because a large proportion of drug
development efforts have recently focused on these settings. A key
objective that we aim to explore with the selected trials is to evaluate
how stage (early, late), line of therapy ({[}neo{]}adjuvant, first line,
advanced lines of therapy), therapy protocols (monotherapy, combination
therapy) and clinical characteristics (all-comer versus complex genetic
or immunological signatures) can be emulated using RWD. If more thorough
feasibility assessments after applying all relevant eligibility criteria
suggest that the threat of bias from mis-measurement of key study
parameters or residual confounding remains high, or that the study size
is not sufficient, then runner-up candidate trials will be considered
instead.

@@

The ENCORE project will use data from four U.S.-based oncology-specific
EHR-derived RWD data sources (in alphabetical order): ConcertAI, COTA,
Flatiron Health, and Ontada/McKesson. A detailed description and
sampling methodology will be provided with each trial emulation
protocol. For ENCORE, not all databases will be available for each
cancer indication given the specificity of data elements required and
the names of the databases will be blinded and referred to as ENCORE
DataBase (EDB) 1, 2, 3 and 4 for the final reporting of results (in
randomly assigned order). If more than one database is considered
fit-for-use for a respective trial emulation, the most suitable analytic
model will be employed for each database separately and final treatment
effect estimates will be pooled using a meta-analytic approach using
fixed effects (weights reflect the inverse of the variance of each
database study) and random effects (weights reflect the inverse of the
variance of each database study plus an additional variance term that
quantifies the assumed heterogeneity between databases)
models.\textsuperscript{20}

@@

For each selected RCT, a detailed protocol, pre-specifying key elements
of the trial emulation, will be developed using the HARPER protocol
template\textsuperscript{21, 22} and will be registered on
ClinicalTrials.gov after review by an expert panel comprising clinicians
and FDA reveiwers. Following the target trial emulation framework, we
will provide an explicit description and rationale on how each element
will be emulated including database selection, covariate measurement,
operationalization of key eligibility criteria, study design, data
analysis and causal contrasts of interest.\textsuperscript{23, 24} Since
it is common that oncology RCTs update survival estimates periodically
based on accrued follow-up time, the protocol will give a brief summary
of each emulated RCT and specify which target OS estimates will be used
for agreement metric evaluation (see
Section~\ref{sec-agreement-metrics}). All eligibility criteria will be
extracted based on publications, publicly available protocols, and
statistical analysis plans of the selected RCT.

@@

An important aspect when emulating oncology trials is the choice and
estimation of the appropriate estimand of interest.\textsuperscript{25}
In prior RCT-DUPLICATE efforts, a ``while on treatment'' strategy was
chosen for database studies that were designed to emulate trials where
the primary trial analysis was intention-to-treat (e.g., a ``treatment
policy'' estimand). This estimand was chosen to mimic the high adherence
typically observed in large cardiovascular clinical trials in the
context of low adherence in clinical practice. However, a ``while on
treatment'' estimand would be inappropriate when emulating oncology
trials, not only because of highly informative censoring, but also
because OS outcome analyses in oncology are typically ``treatment
policy'' estimands that allow for crossover or other post-progression
antineoplastic therapy.\textsuperscript{26} We plan to focus on
``treatment policy'' estimands for the primary analyses in our trial
emulations. However, allowing crossover can dilute treatment effects,
moving point estimates toward the null. Therefore, differences in
cross-over and discontinuation rates between the trial and emulation may
still contribute to differences in observed outcomes. Recognizing these
practical challenges, we will carefully characterize the patterns of
discontinuation or cross-over in both the trials and the database
studies designed to emulate them.

@@

Real-world data fitness and emulation feasibility for each candidate
trial will be assessed in multiple steps based on the approach described
by the Oncology Quality, Characterization, and Assessment of Real-World
Data (Oncology QCARD) Initiative.\textsuperscript{18} The first step
assesses if relevant variables like exposure, line of therapy, outcomes,
and covariates are generally available, measured and operationalizable
in routine-care. Since many oncology RCTs in recent years have focused
on biomarker-defined populations, nuances in measurement and
operationalizability of specific biomarkers must be understood to ensure
a representative and adequately sized study population. For example,
immune checkpoint inhibitors have significantly changed the cancer
treatment landscape since the first approvals of the CTLA-4 inhibitor
ipilimumab in 2011 and the PD-1 inhibitors pembrolizumab and nivolumab
in 2014 in the United States. However, the operationalization of the
expression of the PD-L1 biomarker in RCTs (e.g., percent staining, tumor
proportion score, or combined positive score) has evolved, and as such
PD-L1 '\emph{positivity'} may have different definitions by year based
on different cut-off values.

According to the \emph{Structured Process to Identify Fit-For-Purpose
Data} (SPIFD) framework\textsuperscript{12}, the next step outlines how
eligibility criteria will be ascertained using a color-coded heatmap
that will indicate the level of confidence on how well each criterion
can be emulated in each selected database. As there are eligibility
criteria in oncology clinical trials which either are infeasible to
emulate (e.g., physician-assessed survival prognosis quantification) or
that do not impact clinical care for the emulation of the trial (e.g.,
male patients should be willing to use barrier contraception), the study
team will determine key eligibility criteria for the emulation of the
trial based on consensus.

Additionally, we will provide conceptual and operationalized definitions
on how exposures, outcomes and covariates are defined in each respective
database. There will be special emphasis on how exposure, in context of
disease and line of therapy settings, and the survival outcomes are
emulated. For all considered databases, real-world OS (rwOS) is
typically a composite measure that, depending on the underlying
database, can be derived from different mortality sources (i.e., EHR
documentation, Social Security Death Index, and other linkages). Given
that not all relevant sources that provide mortality data are
synchronized and updated uniformly, sensitivity analyses with more
conservative (i.e., earlier) censoring dates will be considered for each
trial emulation to mitigate the potential impact of ghost-time
bias.\textsuperscript{27}

@@

Particularly when emulating pivotal trials of practice-changing
treatments, multiple aspects need to be considered, such as the
contemporaneity of the control cohort, the adoption rate of the novel
medical product in routine care, the magnitude of the clinical treatment
benefit, and the rate in which patients discontinue or cross-over, as
they could influence effect estimates.

To that end, comprehensive data explorations will be performed as part
of the protocol development to contextualize these parameters and (if
reported) evaluate comparisons to the emulated trial. Examples for such
standard diagnostics are visualized in Figure~\ref{fig-initiators}.

The distribution of patient characteristics, by exposure status, will be
examined before and after applying eligibility criteria and contrasted
with the distributions of patient characteristics of the original RCT.
Initial propensity score matching or weighting methods will be applied
to ensure that measured pre-exposure covariates can be balanced,
exposure cohorts are conditionally exchangeable at baseline, and
resulting sample sizes are still sufficient after matching or weighting.
At this stage, all exploratory analyses will be conducted blinded
towards the outcome to not bias any study design or analytic choices
based on known outcome information.
